---
title: "Probability and Statistics III"
author: "Prof. Josphat Kinyanjui"
date: "May 27, 2026"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
    subtitle: Lecture Notes
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
  word_document:
    toc: true
    toc_depth: '3'
  rmdformats::readthedown:
    highlight: tango
    use_bookdown: true
    toc_depth: 3
    number_sections: true
institute: Department of Mathematics, Statistics & Actuarial Science, Karatina University,
  Kenya
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(tinytex)
```

# Multivariate Distribution Theory

We earlier defined a random variable as a real-valued function over a sample space with a probability measure. Many different random variables can be defined over one and the same sample space.

In this section we shall be concerned first with the **bivariate case**, i.e. situations where we are interested at the same time in a pair of random variables defined over a joint sample space. Later we shall extend this discussion to the **multivariate case**, covering any finite number of random variables.

If $X$ and $Y$ are discrete random variables, we write the probability that $X$ will take on the value $x$ and $Y$ will take on the value $y$ as $f(x, y)$. Thus, $f(x, y)$ is the probability of the intersection of the events $X = x$ and $Y = y$.

## Joint Probability Distribution (Discrete Case)

### Example 1

Two caplets are selected at random from a bottle containing three aspirins, two sedatives, and four laxative caplets.

If $X$ and $Y$ are, respectively, the numbers of aspirin and sedative caplets included among the two caplets drawn, find the probabilities associated with all possible pairs of values of $X$ and $Y$.

**Solution:**

The possible pairs are $(x, y)$ where $x \in \{0, 1, 2\}$ and $y \in \{0, 1, 2\}$, subject to $x + y \leq 2$.

For example, the probability associated with $(1, 0)$: we want one aspirin, zero sedatives, and hence one laxative. The number of ways is $\binom{3}{1}\binom{2}{0}\binom{4}{1}$ and the total number of ways to select two caplets from nine is $\binom{9}{2}$.

Continuing this way, we obtain:

| | **x = 0** | **x = 1** | **x = 2** |
|---|---|---|---|
| **y = 0** | $f(0,0)$ | $f(1,0)$ | $f(2,0)$ |
| **y = 1** | $f(0,1)$ | $f(1,1)$ | 0 |
| **y = 2** | $f(0,2)$ | 0 | 0 |

It is generally preferable to represent such probabilities by means of a formula — a function $f(x, y)$ defined for any pair of values within the range of $X$ and $Y$.

### Definition 1 – Joint Probability Distribution Function (Discrete)

If $X$ and $Y$ are discrete random variables, the function given by $f(x, y) = P(X = x, Y = y)$ for each pair of values $(x, y)$ within the range of $X$ and $Y$ is called the **joint probability distribution function** of $X$ and $Y$.

### Theorem 1

A bivariate function can serve as the joint probability distribution function of a pair of discrete random variables $X$ and $Y$ if and only if its values $f(x, y)$ satisfy the conditions:

1. $f(x, y) \geq 0$, for each pair of values $(x, y)$ within its domain.
2. $\displaystyle\sum_x \sum_y f(x, y) = 1$, where the double summation extends over all possible pairs within the domain.

### Example 2

Determine the value of $k$ for which the function $f(x, y) = k(x + y)$ for $x = 0, 1, 2$ and $y = 0, 1, 2$ can serve as a joint probability distribution function.

**Solution:**

Substituting the various values of $x$ and $y$ and applying Theorem 1, we get:

$$\sum_x \sum_y k(x + y) = 1$$

Solving for $k$ gives the appropriate value.

## Joint Distribution Function (Discrete Case)

### Definition 2 – Joint CDF (Discrete)

If $X$ and $Y$ are discrete random variables, the function given by

$$F(x, y) = P(X \leq x, Y \leq y) = \sum_{s \leq x} \sum_{t \leq y} f(s, t)$$

for $-\infty < x, y < \infty$, where $f(s, t)$ is the value of the joint probability distribution of $X$ and $Y$ at $(s, t)$, is called the **joint distribution function** or the **joint Cumulative Distribution Function (CDF)** of $X$ and $Y$.

### Example 3

With reference to Example 1, find $F(1, 1)$.

**Solution:**

As in the univariate case, the joint distribution function is defined for all real numbers. For example, $F(1, 1) = \sum_{x \leq 1} \sum_{y \leq 1} f(x, y)$.

## Joint Probability Density (Continuous Case)

### Definition 3 – Joint PDF (Continuous)

A bivariate function with values $f(x, y)$, defined over the $xy$-plane, is called a **joint probability density function** of a continuous random variable $X$ and $Y$ if and only if

$$P[(X, Y) \in A] = \iint_A f(x, y) \, dx \, dy$$

for any region $A$ in the $xy$-plane.

### Theorem 2

A bivariate function can serve as a joint probability density function of a pair of continuous random variables $X$ and $Y$ if and only if its values $f(x, y)$ satisfy the conditions:

1. $f(x, y) \geq 0$ for all $(x, y)$
2. $\displaystyle\int_{-\infty}^{\infty}\int_{-\infty}^{\infty} f(x, y) \, dx \, dy = 1$

### Example 4

Given the joint probability density function $f(x, y)$ of two random variables $X$ and $Y$, find $P[(X, Y) \in A]$ where $A$ is a given region.

**Solution:**

$$P[(X, Y) \in A] = \iint_A f(x, y) \, dx \, dy$$

## Joint CDF (Continuous Case)

### Definition 4 – Joint CDF (Continuous)

If $X$ and $Y$ are continuous random variables, the function given by

$$F(x, y) = \int_{-\infty}^{x}\int_{-\infty}^{y} f(s, t) \, dt \, ds, \quad -\infty < x, y < \infty$$

where $f(s, t)$ is the value of the joint probability density of $X$ and $Y$ at $(s, t)$, is called the **joint Cumulative Density Function** of $X$ and $Y$.

Analogous to the univariate case, partial differentiation yields:

$$\frac{\partial^2 F(x, y)}{\partial x \, \partial y} = f(x, y)$$

wherever these partial derivatives exist.

### Example 5

If the joint probability density of $X$ and $Y$ is given by $f(x, y)$, find the cumulative density function.

**Solution:**

$$F(x, y) = \int_{-\infty}^{x}\int_{-\infty}^{y} f(s, t) \, dt \, ds$$

### Example 6

Find the joint probability density of the two random variables $X$ and $Y$ whose joint cumulative density function is given. Also use the joint probability density function to determine the required probability.

**Solution:**

(a) Since $f(x, y) = \dfrac{\partial^2 F(x,y)}{\partial x\, \partial y}$, partial differentiation yields the joint PDF.

(b) The probability is evaluated by integrating $f(x,y)$ over the relevant region.

For two random variables, the joint probability density is geometrically a surface, and the probability is given by the volume under this surface over the region $A$.

## Multivariate Case

All the definitions of this section can be generalized to the **multivariate case** with $n$ random variables. Corresponding to Definition 1, the values of the joint probability distribution of $n$ discrete random variables $X_1, X_2, \ldots, X_n$ are given by

$$f(x_1, x_2, \ldots, x_n) = P(X_1 = x_1, X_2 = x_2, \ldots, X_n = x_n)$$

for each $n$-tuple $(x_1, \ldots, x_n)$ within the range of the random variables; and their joint distribution function is

$$F(x_1, \ldots, x_n) = P(X_1 \leq x_1, \ldots, X_n \leq x_n)$$

### Example 7

If the joint probability distribution of three random variables $X$, $Y$, and $Z$ is given, find the specified probability.

**Solution:**

$$P(X = x, Y = y, Z = z) = \sum_{x}\sum_{y}\sum_{z} f(x, y, z)$$

In the continuous case, probabilities are obtained by integrating the joint probability density, and the joint CDF is given by

$$F(x_1, \ldots, x_n) = \int_{-\infty}^{x_1} \cdots \int_{-\infty}^{x_n} f(t_1, \ldots, t_n) \, dt_n \cdots dt_1$$

### Example 8

If the trivariate probability density of $(X, Y, Z)$ is given by $f(x, y, z)$, find the probability over a specified region $A$.

**Solution:**

$$P[(X, Y, Z) \in A] = \iiint_A f(x, y, z) \, dz \, dy \, dx$$

---

##Exercises

1. Given the joint probability distribution table of $X$ and $Y$ (with rows $y = 0, 1, 2, 3$ and columns $x = 0, 1, 2$), find the specified probabilities.

2. If the joint probability density of $X$ and $Y$ is given, find the value of $c$, then find:
   (a) $P(X < 1, Y < 1)$
   (b) $P(X + Y < 1)$
   (c) $P(X > Y)$

3. Determine $k$ so that a given function can serve as a joint probability density function.

4. Find $F(x, y)$ if the joint probability distribution of $X$ and $Y$ is given.

5. Find $k$ if a given bivariate function can serve as a joint probability density.

6. Verify the joint distribution function of a given example.

7. If a trivariate probability density $f(x_1, x_2, x_3)$ is given, find the specified probability.

8. A certain college gives attitude tests in the sciences and humanities. If $X$ and $Y$ are the proportions of correct answers in the two subjects, find the probabilities that a student will get:
   (a) Less than $\frac{1}{2}$ on both tests.
   (b) More than $\frac{1}{2}$ on the science test and less than $\frac{1}{4}$ on the humanities test.

9. Suppose $P$, the price of a commodity (in dollars), and $S$, its total sales (in units), are random variables with a given joint probability density. Find the probabilities that:
   (a) The price will be less than 30 cents and sales will exceed 10,000 units.
   (b) The price will be between 25 cents and 30 cents and sales will be less than 6,000 units.

---

# Marginal and Conditional Distributions

## Marginal Distributions

### Definition 5 – Marginal Distribution (Discrete)

If $X$ and $Y$ are discrete random variables and $f(x, y)$ is the value of their joint probability distribution at $(x, y)$, the function given by

$$g(x) = \sum_{y} f(x, y), \quad \text{for each } x \text{ within the range of } X$$

is called the **marginal probability distribution function** of $X$.

Correspondingly, the function given by

$$h(y) = \sum_{x} f(x, y), \quad \text{for each } y \text{ within the range of } Y$$

is called the **marginal probability distribution function** of $Y$.

### Example 9

The joint probability distribution function of random variables $X$ and $Y$ is given in the table below ($x, y \in \{1, 2, 3\}$). Determine the marginal probability distribution functions of $X$ and $Y$.

| | **x = 1** | **x = 2** | **x = 3** | $h(y)$ |
|---|---|---|---|---|
| **y = 1** | $f(1,1)$ | $f(2,1)$ | $f(3,1)$ | |
| **y = 2** | $f(1,2)$ | $f(2,2)$ | $f(3,2)$ | |
| **y = 3** | $f(1,3)$ | $f(2,3)$ | $f(3,3)$ | |
| $g(x)$ | | | | |

**Solution:**

(i) The marginal probability distribution function of $X$: $g(x) = \sum_y f(x, y)$

(ii) The marginal probability distribution function of $Y$: $h(y) = \sum_x f(x, y)$

### Definition 6 – Marginal Density (Continuous)

If $X$ and $Y$ are continuous random variables and $f(x, y)$ is the value of their joint probability density at $(x, y)$, the function given by

$$g(x) = \int_{-\infty}^{\infty} f(x, y) \, dy$$

is called the **marginal probability density function** of $X$.

Correspondingly,

$$h(y) = \int_{-\infty}^{\infty} f(x, y) \, dx$$

is called the **marginal probability density function** of $Y$.

### Example 10

Given the joint probability density $f(x, y)$, find the marginal probability density functions of $X$ and $Y$.

**Solution:**

$$g(x) = \int_{-\infty}^{\infty} f(x, y) \, dy, \qquad h(y) = \int_{-\infty}^{\infty} f(x, y) \, dx$$

## Joint Marginal Distributions (Multivariate)

When dealing with more than two random variables, we can speak of **joint marginal distributions** of several of the random variables.

If the joint probability of discrete random variables $X_1, X_2, \ldots, X_n$ has the value $f(x_1, x_2, \ldots, x_n)$, the marginal probability distribution function of $X_1$ alone is:

$$f_1(x_1) = \sum_{x_2} \cdots \sum_{x_n} f(x_1, x_2, \ldots, x_n)$$

The joint marginal probability distribution function of $X_1$ and $X_2$ is:

$$f_{12}(x_1, x_2) = \sum_{x_3} \cdots \sum_{x_n} f(x_1, x_2, \ldots, x_n)$$

For the **continuous case**, summations are replaced by integrals:

$$f_1(x_1) = \int_{-\infty}^{\infty} \cdots \int_{-\infty}^{\infty} f(x_1, \ldots, x_n) \, dx_2 \cdots dx_n$$

### Example 11

Considering the trivariate probability density $f(x_1, x_2, x_3)$, find:

(i) The joint marginal probability density function of $X_1$ and $X_2$.
(ii) The marginal probability density function of $X_1$ alone.

**Solution:**

(i) $f_{12}(x_1, x_2) = \int_{-\infty}^{\infty} f(x_1, x_2, x_3) \, dx_3$

(ii) $f_1(x_1) = \int_{-\infty}^{\infty} f_{12}(x_1, x_2) \, dx_2$

---

## Conditional Distributions

We define the conditional probability of event $A$ given event $B$ as

$$P(A \mid B) = \frac{P(A \cap B)}{P(B)}, \quad \text{provided } P(B) > 0$$

Suppose $A$ is the event $\{X = x\}$ and $B$ is the event $\{Y = y\}$. Then:

$$P(X = x \mid Y = y) = \frac{f(x, y)}{h(y)}, \quad \text{provided } h(y) > 0$$

### Definition 7 – Conditional Distribution (Discrete)

If $f(x, y)$ is the value of the joint probability distribution of the discrete random variables $X$ and $Y$ at $(x, y)$, and $h(y)$ is the value of the marginal distribution of $Y$ at $y$, the function given by

$$f(x \mid y) = \frac{f(x, y)}{h(y)}, \quad \text{for each } x \text{ within the range of } X$$

is called the **conditional distribution of $X$ given $Y = y$**.

Correspondingly, if $g(x)$ is the value of the marginal distribution of $X$ at $x$:

$$f(y \mid x) = \frac{f(x, y)}{g(x)}, \quad \text{for each } y \text{ within the range of } Y$$

is called the **conditional distribution of $Y$ given $X = x$**.

### Definition 8 – Conditional Density (Continuous)

If $f(x, y)$ is the value of the joint density of continuous random variables $X$ and $Y$ at $(x, y)$, and $h(y)$ is the value of the marginal density of $Y$ at $y$, the function given by

$$f(x \mid y) = \frac{f(x, y)}{h(y)}, \quad \text{for } h(y) > 0$$

is called the **conditional density of $X$ given $Y = y$**.

Correspondingly:

$$f(y \mid x) = \frac{f(x, y)}{g(x)}, \quad \text{for } g(x) > 0$$

is called the **conditional density of $Y$ given $X = x$**.

### Example 12

Given the joint probability density $f(x, y)$, find the marginal densities of $X$ and $Y$ and the conditional density of $X$ given $Y = y$.

**Solution:**

$$g(x) = \int f(x, y) \, dy, \qquad h(y) = \int f(x, y) \, dx$$

$$f(x \mid y) = \frac{f(x, y)}{h(y)}$$

### Example 13

With reference to Example 1, find:

(i) The conditional distribution of $X$ given $Y = 1$.
(ii) The conditional distribution of $Y$ given $X = 0$.

**Solution:**

The joint distribution together with marginal totals:

| | **x = 0** | **x = 1** | **x = 2** | $h(y)$ |
|---|---|---|---|---|
| **y = 0** | $f(0,0)$ | $f(1,0)$ | $f(2,0)$ | |
| **y = 1** | $f(0,1)$ | $f(1,1)$ | 0 | |
| **y = 2** | $f(0,2)$ | 0 | 0 | |
| $g(x)$ | | | | |

(i) Conditional distribution of $X$ given $Y = 1$: $\quad f(x \mid 1) = \dfrac{f(x, 1)}{h(1)}$

(ii) Conditional distribution of $Y$ given $X = 0$: $\quad f(y \mid 0) = \dfrac{f(0, y)}{g(0)}$

---

## Independence of Random Variables

When we deal with two or more random variables, questions of **independence** are usually of great importance.

If the values of the conditional distribution of $X$ given $Y = y$ do not depend on $y$, then $f(x \mid y) = g(x)$, and the formula yields:

$$f(x, y) = g(x) \cdot h(y)$$

That is, the values of the joint distribution are given by the products of the corresponding marginal distributions.

### Definition 9 – Independence (Discrete)

If $f(x_1, x_2, \ldots, x_n)$ is the value of the joint probability distribution of $n$ discrete random variables $X_1, \ldots, X_n$ at $(x_1, \ldots, x_n)$, and $f_i(x_i)$ is the value of the marginal distribution of $X_i$, then the $n$ random variables are **independent** if and only if

$$f(x_1, \ldots, x_n) = f_1(x_1) \cdot f_2(x_2) \cdots f_n(x_n)$$

for all $(x_1, \ldots, x_n)$ within their range.

### Definition 10 – Independence (Continuous)

If $f(x_1, x_2, \ldots, x_n)$ is the value of the joint probability density function of $n$ continuous random variables $X_1, \ldots, X_n$ at $(x_1, \ldots, x_n)$, and $f_i(x_i)$ is the value of the marginal density of $X_i$, then the $n$ random variables are **independent** if and only if

$$f(x_1, \ldots, x_n) = f_1(x_1) \cdot f_2(x_2) \cdots f_n(x_n)$$

for all $(x_1, \ldots, x_n)$ within their range.

### Example 14

If the joint probability density of $X$ and $Y$ is given, find:

(a) The marginal density of $X$.
(b) The marginal density of $Y$.
(c) Whether the two random variables are independent.

**Solution:**

Compute $g(x) = \int f(x,y)\,dy$ and $h(y) = \int f(x,y)\,dx$. Then check whether $f(x,y) = g(x) \cdot h(y)$.

If $f(x,y) \neq g(x) \cdot h(y)$, the two random variables are **not independent**.

### Example 15

With reference to Example 1, determine whether $X$ and $Y$ are independent.

**Solution:**

Using the marginal distributions obtained previously, check whether $f(x,y) = g(x) \cdot h(y)$ for all pairs. If any pair fails, $X$ and $Y$ are **not independent**.

### Example 16

Considering independent flips of a balanced coin, let $X_i$ be the number of heads (0 or 1) obtained on the $i$-th flip. Find the joint probability distribution of these $n$ random variables.

**Solution:**

Since each $X_i$ has the distribution $P(X_i = x) = \frac{1}{2}$ for $x = 0, 1$, and the random variables are independent, their joint probability distribution is:

$$f(x_1, \ldots, x_n) = \prod_{i=1}^{n} \frac{1}{2} = \left(\frac{1}{2}\right)^n, \quad x_i \in \{0, 1\}$$

### Example 17

Given independent random variables $X_1$, $X_2$, $X_3$ with probability densities $f_1(x_1)$, $f_2(x_2)$, $f_3(x_3)$, find their joint probability density and use it to evaluate a specified probability.

**Solution:**

The joint probability density function is:

$$f(x_1, x_2, x_3) = f_1(x_1) \cdot f_2(x_2) \cdot f_3(x_3)$$

---

## Exercises

1. Given the joint probability distribution table of $X$ and $Y$ (with $x \in \{-1, 1\}$ and $y \in \{-1, 0, 1\}$), find:
   (a) Marginal distribution of $X$.
   (b) Marginal distribution of $Y$.
   (c) Conditional distribution of $X$ given $Y = 1$.
   (d) Conditional distribution of $Y$ given $X = 0$.

2. Given the joint probability distribution $f(x, y, z)$, find:
   - The joint marginal distribution of $X$ and $Y$.
   - The joint marginal distribution of $X$ and $Z$.
   - The marginal distribution of $X$.
   - The conditional distribution of $Z$ given $X = x$ and $Y = y$.
   - The joint conditional distribution of $Y$ and $Z$ given $X = x$.

3. Given the joint probability distribution table of $X$ and $Y$, find:
   (a) Marginal distribution of $X$.
   (b) Marginal distribution of $Y$.
   (c) Conditional distribution of $X$ given $Y = -1$.

4. Check whether $X$ and $Y$ are independent for given joint distributions.

5. If the joint probability density of $X$ and $Y$ is given, find:
   - The marginal density of $X$.
   - The conditional density of $Y$ given $X = x$.

6. If $X$ is the proportion of persons responding to one kind of mail-order solicitation and $Y$ to another, find the probabilities that:
   - At least 30% will respond to the first kind.
   - At most 50% will respond to the second kind given a 20% response to the first.

7. If $X$ is the money (in dollars) a salesperson spends on gasoline per day and $Y$ is the corresponding reimbursement, find:
   - The marginal density of $X$.
   - The conditional density of $Y$ given $X = x$.
   - The probability that the salesperson will be reimbursed at least \$8 when spending \$12.

8. The useful life (in hours) of a vacuum tube is a random variable with probability density $f(x)$. If three tubes operate independently, find:
   (a) The joint probability density of their lifetimes $(X_1, X_2, X_3)$.
   (b) The value of $P(X_1 < X_2 < X_3)$ (or as specified).

---

# The Bivariate Normal Distribution

Among multivariate densities of special importance is the **multivariate normal distribution**, which is a generalization of the normal distribution in one variable.

## Definition and Properties

### Definition 1 – Bivariate Normal Distribution

A pair of random variables $X$ and $Y$ have a **bivariate normal distribution** and are referred to as **jointly normally distributed** random variables, if and only if their joint probability density is given by:

$$f(x, y) = \frac{1}{2\pi\sigma_1\sigma_2\sqrt{1-\rho^2}} \exp\!\left\{-\frac{1}{2(1-\rho^2)}\left[\frac{(x-\mu_1)^2}{\sigma_1^2} - \frac{2\rho(x-\mu_1)(y-\mu_2)}{\sigma_1\sigma_2} + \frac{(y-\mu_2)^2}{\sigma_2^2}\right]\right\}$$

where $-1 < \rho < 1$.

To study this joint distribution, we first show that the parameters $\mu_1, \mu_2$ are the means and $\sigma_1, \sigma_2$ are the standard deviations of $X$ and $Y$ respectively.

Integrating on $y$ from $-\infty$ to $\infty$ gives the **marginal density of $X$**. After completing the square and simplifying:

$$g(x) = \frac{1}{\sigma_1\sqrt{2\pi}} e^{-\frac{(x-\mu_1)^2}{2\sigma_1^2}}$$

which is a **normal distribution** with mean $\mu_1$ and standard deviation $\sigma_1$. By symmetry, the marginal density of $Y$ is a normal distribution with mean $\mu_2$ and standard deviation $\sigma_2$.

The parameter $\rho$ is called the **correlation coefficient**, and it can be shown that $-1 < \rho < 1$. It measures how the two random variables $X$ and $Y$ vary together.

## Conditional Distributions

### Theorem 1 – Conditional Distributions of the Bivariate Normal

If $X$ and $Y$ have a bivariate normal distribution, the **conditional density of $Y$ given $X = x$** is a normal distribution with:

- **Mean:** $\mu_2 + \rho\dfrac{\sigma_2}{\sigma_1}(x - \mu_1)$
- **Variance:** $\sigma_2^2(1 - \rho^2)$

And the **conditional density of $X$ given $Y = y$** is a normal distribution with:

- **Mean:** $\mu_1 + \rho\dfrac{\sigma_1}{\sigma_2}(y - \mu_2)$
- **Variance:** $\sigma_1^2(1 - \rho^2)$

**Proof:**

Writing $f(y \mid x) = \dfrac{f(x, y)}{g(x)}$ and simplifying the exponent, one arrives at a normal density with the stated mean and variance. The corresponding result for the conditional density of $X$ given $Y = y$ follows by symmetry.

## Key Theorems

### Theorem 2 – Independence in the Bivariate Normal

If two random variables have a bivariate normal distribution, they are **independent if and only if $\rho = 0$**.

**Proof:** (exercise)

> **Remark:** When $\rho = 0$, the random variables are said to be **uncorrelated**. Note that the marginal distributions may both be normal without the joint distribution being bivariate normal — the converse does not necessarily hold.

## Examples

### Example 1

In a certain population of married couples, the height $X$ (in meters) of the husband and the height $Y$ of the wife have a bivariate normal distribution with given parameters.

(i) Find the mean and the variance of the height for the husbands and wives.
(ii) Find the expected height and variance of the husband whose wife is expected to be 1.55 m tall.

**Solution:**

(i) Comparing the given density with the general formula, we identify:

$$\mu_1 = \ldots, \quad \sigma_1 = \ldots, \quad \mu_2 = \ldots, \quad \sigma_2 = \ldots, \quad \rho = \ldots$$

(ii) By Theorem 1, the conditional mean of $X$ given $Y = 1.55$ is:

$$E(X \mid Y = 1.55) = \mu_1 + \rho\frac{\sigma_1}{\sigma_2}(1.55 - \mu_2)$$

The conditional variance of $X$ given $Y = y$ is:

$$\text{Var}(X \mid Y = y) = \sigma_1^2(1 - \rho^2)$$

### Example 2

Let $X$ and $Y$ have a bivariate normal distribution with parameters $\mu_1, \mu_2, \sigma_1, \sigma_2, \rho$.

Find:

(i) $P(\ldots)$ [2 marks]
(ii) $P(\ldots \mid X = 74)$ [6 marks]

**Solution:**

(i) Compute using the standard normal CDF.

(ii) The conditional distribution of $Y$ given $X = 74$ is normal with:

$$\text{Mean} = \mu_2 + \rho\frac{\sigma_2}{\sigma_1}(74 - \mu_1), \qquad \text{Variance} = \sigma_2^2(1 - \rho^2)$$

Hence, standardize and use the standard normal table to find the required probability.

---

## Exercises

1. If the exponent of $e$ of a bivariate normal density is given, find:
   (a) The parameters $\mu_1, \mu_2, \sigma_1, \sigma_2, \rho$.
   (b) The required probability.

2. If the exponent of $e$ of a bivariate normal density is given, find $P(Y < y_0 \mid X = x_0)$.

3. If $X$ and $Y$ have a bivariate normal distribution with specified parameters, find $P(\ldots)$.

4. If $U$ and $V$ have a bivariate normal distribution, and $U = aX + b$, $V = cY + d$, find an expression for the correlation coefficient of $U$ and $V$.

5. If $X$ and $Y$ have a bivariate normal distribution, their joint moment-generating function is given by:

$$M(t_1, t_2) = \exp\!\left(\mu_1 t_1 + \mu_2 t_2 + \tfrac{1}{2}\sigma_1^2 t_1^2 + \rho\sigma_1\sigma_2 t_1 t_2 + \tfrac{1}{2}\sigma_2^2 t_2^2\right)$$

Verify that:
   - The first partial derivative with respect to $t_1$ at $t_1 = t_2 = 0$ equals $\mu_1$.
   - The second partial derivative with respect to $t_1^2$ at $t_1 = t_2 = 0$ equals $\mu_1^2 + \sigma_1^2$.
   - The second partial derivative with respect to $t_1$ and $t_2$ at $t_1 = t_2 = 0$ equals $\mu_1\mu_2 + \rho\sigma_1\sigma_2$.

6. Suppose that $X$ and $Y$, the height and weight of certain animals, have a bivariate normal distribution with $\mu_1 = 22$ inches, $\mu_2 = 15$ pounds, $\sigma_1 = 2$ inches, $\sigma_2 = 3$ pounds, and $\rho = 0.80$. Find:
   (a) The expected weight of an animal that is 25 inches tall. **[14.5 pounds]**
   (b) The expected height of an animal that weighs 20 pounds. **[23.625 inches]**
   
# Transformation of Random Variables {#week4}

We concern ourselves with the problem of finding the probability distributions or densities of **functions of one or more random variables**. Given a set of random variables $X_1, X_2, \ldots, X_n$ and their joint probability or density, we are interested in finding the joint probability distribution of some random variable:

$$Y = u(X_1, X_2, \ldots, X_n)$$

This means the values of $Y$ are related to those of the $X$'s by the equation $y = u(x_1, x_2, \ldots, x_n)$.

Several methods are available for solving this kind of problem. The ones we discuss are:

1. The **Distribution Function Technique**
2. The **Transformation Technique**
3. The **Moment-Generating Function Technique**

---

## The Distribution Function Technique {#dist-fn-technique}

A straightforward method of obtaining the probability density of a function of continuous random variables consists of:

1. First finding its **distribution function**, then
2. Finding its **probability density by differentiation**.

Thus, if $X_1, X_2, \ldots, X_n$ are continuous random variables with a given joint probability density, the probability density of $Y = u(X_1, X_2, \ldots, X_n)$ is obtained by first determining:

$$F_Y(y) = \Pr(Y \leq y) = \Pr\!\left[u(X_1, X_2, \ldots, X_n) \leq y\right]$$

and then differentiating to get:

$$f(y) = \frac{d}{dy} F(y)$$

### Example 1 {#ex1}

**Problem:** The probability density of $X$ is given by:

$$f(x) = \begin{cases} 6x(1 - x), & 0 < x < 1 \\ 0, & \text{elsewhere} \end{cases}$$

Find the probability density of $Y = X^3$.

**Solution:**

Letting $G(y)$ denote the distribution function of $Y$:

$$G(y) = \Pr(Y \leq y) = \Pr(X^3 \leq y) = \Pr\!\left(X \leq y^{1/3}\right)$$

$$= \int_0^{y^{1/3}} 6x(1-x)\, dx = 3y^{2/3} - 2y$$

Differentiating:

$$g(y) = \frac{d}{dy} G(y) = 2y^{-1/3} - 2 = \frac{2}{3}\left(y^{-1/3} - 1\right) \cdot 3 = 2\left(y^{-1/3} - 1\right)$$

More precisely:

$$g(y) = \frac{1}{3}\left(2y^{-1/3} - 2\right)$$

$$\boxed{g(y) = \begin{cases} \frac{1}{3}(2y^{-1/3} - 2), & 0 < y < 1 \\ 0, & \text{elsewhere} \end{cases}}$$

### Example 2 {#ex2}

**Problem:** If $Y = |X|$, show that:

$$g(y) = \begin{cases} f(y) + f(-y), & y > 0 \\ 0, & \text{elsewhere} \end{cases}$$

where $f(x)$ is the probability density of $X$.

**Solution:**

For $y > 0$:

$$G(y) = \Pr(Y \leq y) = \Pr(|X| \leq y) = \Pr(-y \leq X \leq y) = F(y) - F(-y)$$

Differentiating:

$$g(y) = \frac{d}{dy}\left[F(y) - F(-y)\right] = f(y) \cdot 1 - f(-y) \cdot (-1) = f(y) + f(-y)$$

Since $X$ cannot be negative, $g(y) = 0$ for $y < 0$. Letting $g(0) = 0$:

$$\boxed{g(y) = \begin{cases} f(y) + f(-y), & y > 0 \\ 0, & \text{elsewhere} \end{cases}}$$

### Example 3 {#ex3}

**Problem:** Given the joint density of $X_1$ and $X_2$:

$$f(x_1, x_2) = \begin{cases} 6e^{-x_1 - 2x_2}, & x_1 > 0,\ x_2 > 0 \\ 0, & \text{elsewhere} \end{cases}$$

Find the probability density of $Y = X_1 + X_2$.

**Solution:**

Integrating the joint density over the region where $x_1 + x_2 \leq y$:

$$F(y) = \int_0^y \int_0^{y - x_1} 6e^{-x_1 - 2x_2}\, dx_2\, dx_1$$

After integration:

$$F(y) = 1 - 2e^{-y} + 3e^{-2y} - 2e^{-3y} \quad \text{(after simplification)}$$

Differentiating gives the density:

$$\boxed{f(y) = \begin{cases} 6e^{-2y} - 6e^{-3y}, & y > 0 \\ 0, & \text{elsewhere} \end{cases}}$$

---

## Exercises — Week 4 {#ex-week4}

1. If $f(x) = 2xe^{-x^2}$ for $x > 0$ (0 elsewhere) and $Y = X^2$, find:
   - (a) The distribution function of $Y$: $G(y) = 1 - e^{-y}$, $y > 0$
   - (b) The probability density of $Y$: $g(y) = e^{-y}$, $y > 0$

2. If $X$ has an exponential distribution with parameter $\lambda$, use the distribution function technique to find the p.d.f. of $Y = \ln X$.

3. If $X$ has the uniform density with $\alpha = 0$ and $\beta = 1$, find the p.d.f. of $Y = \sqrt{X}$:
$$g(y) = \begin{cases} 2y, & 0 < y < 1 \\ 0, & \text{elsewhere} \end{cases}$$

4. If the joint p.d.f. of $X$ and $Y$ is $f(x,y) = 4xye^{-(x^2+y^2)}$ for $x,y > 0$ (0 elsewhere) and $Z = X^2 + Y^2$, find:
   - (a) The distribution function of $Z$
   - (b) The probability density of $Z$

5. If $X_1$ and $X_2$ are independent exponential random variables with parameters $\lambda_1$ and $\lambda_2$, find the p.d.f. of $Y = X_1 + X_2$ when:
   - (a) $\lambda_1 \neq \lambda_2$: $f(y) = \dfrac{\lambda_1\lambda_2}{\lambda_1 - \lambda_2}\left(e^{-\lambda_2 y} - e^{-\lambda_1 y}\right)$, $y > 0$
   - (b) $\lambda_1 = \lambda_2 = \lambda$: $f(y) = \lambda^2 y e^{-\lambda y}$, $y > 0$

6. With reference to Exercise 5 (when $\lambda_1 \neq \lambda_2$), show that $Z = \dfrac{X_1}{X_1 + X_2}$ has a uniform density with $\alpha = 0$, $\beta = 1$.

7. If the joint density of $X$ and $Y$ is $f(x,y) = e^{-(x+y)}$ for $x,y > 0$ (0 elsewhere) and $Z = \dfrac{X+Y}{2}$, find the p.d.f. of $Z$ using the distribution function technique.

8. The percentages of copper and iron in a certain ore are $X_1$ and $X_2$ respectively. Their joint density is:
$$f(x_1, x_2) = \begin{cases} 5, & 0 \leq x_1 \leq 1,\ 0 \leq x_2 \leq \frac{2-x_1}{2} \\ 0, & \text{elsewhere} \end{cases}$$
Find the p.d.f. of $Y = X_1 + X_2$ and $E(Y)$.

$$g(y) = \begin{cases} \tfrac{11}{9}\cdot\tfrac{y}{2}, & 0 \leq y < 1 \\ \tfrac{3(2-y)(7-4y)}{22}, & 1 \leq y \leq 2 \\ 0, & \text{elsewhere} \end{cases}$$

---

# Transformation Technique {#week5}

## One Variable {#one-var}

We can determine the probability distribution or density of a function of a random variable **without first finding its distribution function**.

### Discrete Case

In the discrete case, when the relationship between the values of $X$ and $Y = u(X)$ is **one-to-one**, we simply make the appropriate substitution.

### Example 4 {#ex4}

**Problem:** If $X$ is the number of heads in four tosses of a fair coin, find the probability distribution of:

$$Y = \frac{1}{X + 1}$$

**Solution:**

Using the binomial distribution with $n = 4$, $p = \frac{1}{2}$:

| $x$ | 0 | 1 | 2 | 3 | 4 |
|-----|---|---|---|---|---|
| $f(x)$ | $\frac{1}{16}$ | $\frac{4}{16}$ | $\frac{6}{16}$ | $\frac{4}{16}$ | $\frac{1}{16}$ |

Applying $y = \frac{1}{x+1}$:

| $y$ | 1 | $\frac{1}{2}$ | $\frac{1}{3}$ | $\frac{1}{4}$ | $\frac{1}{5}$ |
|-----|---|---|---|---|---|
| $g(y)$ | $\frac{1}{16}$ | $\frac{4}{16}$ | $\frac{6}{16}$ | $\frac{4}{16}$ | $\frac{1}{16}$ |

Substituting $x = \frac{1}{y} - 1$ into the binomial formula directly:

$$g(y) = \binom{4}{\frac{1}{y}-1}\left(\frac{1}{2}\right)^4, \quad y = 1, \tfrac{1}{2}, \tfrac{1}{3}, \tfrac{1}{4}, \tfrac{1}{5}$$

> **Note:** The probabilities remain unchanged — only the variable changes from $X$ to $Y$.

### Example 5 — Non One-to-One Case {#ex5}

**Problem:** Find the probability distribution of $Z = (X - 2)^2$ where $X$ is as in Example 4.

**Solution:**

| $x$ | 0 | 1 | 2 | 3 | 4 |
|-----|---|---|---|---|---|
| $(x-2)^2 = z$ | 4 | 1 | 0 | 1 | 4 |

Solving $x = 2 \pm \sqrt{z}$ and summing probabilities for each $z$:

$$h(0) = f(2) = \frac{6}{16} = \frac{3}{8}$$

$$h(1) = f(1) + f(3) = \frac{4}{16} + \frac{4}{16} = \frac{8}{16} = \frac{1}{2}$$

$$h(4) = f(0) + f(4) = \frac{1}{16} + \frac{1}{16} = \frac{2}{16} = \frac{1}{8}$$

| $z$ | 0 | 1 | 4 |
|-----|---|---|---|
| $h(z)$ | $\frac{3}{8}$ | $\frac{1}{2}$ | $\frac{1}{8}$ |

### Continuous Case — Theorem 1 {#theorem1}

> **Theorem 1:** Let $f(x)$ be the probability density of the continuous random variable $X$. If $y = u(x)$ is differentiable and either **strictly increasing or strictly decreasing** for all values where $f(x) > 0$, then the equation $y = u(x)$ can be uniquely solved for $x = w(y)$, and the probability density of $Y = u(X)$ is:
>
> $$g(y) = f[w(y)] \cdot \left|w'(y)\right|, \quad \text{provided } u'(x) \neq 0$$
>
> and $g(y) = 0$ elsewhere.

**Proof sketch:** For the increasing case, $\Pr(a < Y < b) = \Pr(w(a) < X < w(b))$. Changing variables $x = w(y)$, $dx = w'(y)\,dy$ in the integral yields $g(y) = f[w(y)] \cdot |w'(y)|$. The decreasing case follows similarly, with the absolute value ensuring positivity. $\square$

### Example 6 {#ex6}

**Problem:** If $X$ has the exponential distribution $f(x) = e^{-x}$ for $x > 0$ (0 elsewhere), find the p.d.f. of $Y = \sqrt{X}$.

**Solution:**

The inverse is $x = y^2$, so $w(y) = y^2$ and $w'(y) = 2y$.

$$g(y) = f(y^2) \cdot |2y| = e^{-y^2} \cdot 2y = 2ye^{-y^2}$$

$$\boxed{g(y) = \begin{cases} 2ye^{-y^2}, & y > 0 \\ 0, & \text{elsewhere} \end{cases}}$$

### Example 7 — Cauchy Distribution {#ex7}

**Problem:** A spinner gives angle $\Theta$ with uniform density $f(\theta) = \frac{1}{2\pi}$ for $-\frac{\pi}{2} < \theta < \frac{\pi}{2}$ (0 elsewhere). Find the density of $X = a\tan\Theta$, the abscissa on the $x$-axis.

**Solution:**

The inverse is $\theta = \arctan(x/a)$, so:

$$\frac{d\theta}{dx} = \frac{a}{a^2 + x^2}$$

$$g(x) = \frac{1}{2\pi} \cdot \frac{a}{a^2 + x^2} \cdot 2\pi = \frac{a}{\pi(a^2 + x^2)} \quad \text{for } -\infty < x < \infty$$

This is the **Cauchy distribution**.

---

## Several Variables {#several-vars}

### Motivation {#motivation}

Given the joint distribution of $X_1$ and $X_2$, we want the distribution of $Y = u(X_1, X_2)$.

**Strategy:** Introduce an auxiliary variable $Z$ (e.g., $Z = X_2$), find the joint density of $(Y, Z)$, then integrate out $Z$ to obtain the marginal density of $Y$.

### Theorem 2 — Jacobian Transformation {#theorem2}

> **Theorem 2:** Let $f(x_1, x_2)$ be the joint probability density of the continuous random variables $X_1$ and $X_2$. If:
>
> $$y_1 = u_1(x_1, x_2) \quad \text{and} \quad y_2 = u_2(x_1, x_2)$$
>
> are partially differentiable and define a **one-to-one transformation**, with inverse $x_i = w_i(y_1, y_2)$, then the joint density of $Y_1 = u_1(X_1, X_2)$ and $Y_2 = u_2(X_1, X_2)$ is:
>
> $$g(y_1, y_2) = f[w_1(y_1, y_2),\ w_2(y_1, y_2)] \cdot |J|$$
>
> where the **Jacobian** is:
>
> $$J = \begin{vmatrix} \dfrac{\partial x_1}{\partial y_1} & \dfrac{\partial x_1}{\partial y_2} \\[8pt] \dfrac{\partial x_2}{\partial y_1} & \dfrac{\partial x_2}{\partial y_2} \end{vmatrix}$$

### Example 8 — Poisson Sum {#ex8}

**Problem:** $X_1$ and $X_2$ are independent Poisson random variables with parameters $\lambda_1$ and $\lambda_2$. Find the distribution of $Y = X_1 + X_2$.

**Solution:**

Their joint distribution is:

$$f(x_1, x_2) = \frac{e^{-\lambda_1}\lambda_1^{x_1}}{x_1!} \cdot \frac{e^{-\lambda_2}\lambda_2^{x_2}}{x_2!}$$

Setting $y = x_1 + x_2$ (so $x_1 = y - x_2$) and summing over $x_2 = 0, 1, \ldots, y$:

$$h(y) = \sum_{x_2=0}^{y} \frac{e^{-(\lambda_1+\lambda_2)}\lambda_1^{y-x_2}\lambda_2^{x_2}}{(y-x_2)!\, x_2!} = \frac{e^{-(\lambda_1+\lambda_2)}(\lambda_1+\lambda_2)^y}{y!}$$

using the binomial expansion $(\lambda_1 + \lambda_2)^y = \sum_{x_2=0}^{y}\binom{y}{x_2}\lambda_1^{y-x_2}\lambda_2^{x_2}$.

> **Conclusion:** The sum of two independent Poisson random variables with parameters $\lambda_1$ and $\lambda_2$ is Poisson with parameter $\lambda_1 + \lambda_2$.

### Example 9 — Exponential to Beta/Gamma {#ex9}

**Problem:** Given joint density $f(x_1, x_2) = e^{-(x_1+x_2)}$ for $x_1, x_2 > 0$ (0 elsewhere), find:

- (a) The joint density of $Y_1 = X_1 + X_2$ and $Y_2 = \dfrac{X_1}{X_1 + X_2}$
- (b) The marginal density of $Y_2$

**Solution (a):**

Solving: $x_1 = y_1 y_2$ and $x_2 = y_1(1 - y_2)$.

$$J = \begin{vmatrix} y_2 & y_1 \\ 1-y_2 & -y_1 \end{vmatrix} = -y_1 y_2 - y_1(1-y_2) = -y_1$$

$$|J| = y_1$$

$$g(y_1, y_2) = e^{-y_1 y_2 - y_1(1-y_2)} \cdot y_1 = y_1 e^{-y_1}$$

$$\boxed{g(y_1, y_2) = \begin{cases} y_1 e^{-y_1}, & y_1 > 0,\ 0 < y_2 < 1 \\ 0, & \text{elsewhere} \end{cases}}$$

**Solution (b):**

$$h(y_2) = \int_0^\infty y_1 e^{-y_1}\, dy_1 = \Gamma(2) = 1! = 1$$

$$\boxed{h(y_2) = \begin{cases} 1, & 0 < y_2 < 1 \\ 0, & \text{elsewhere} \end{cases}}$$

This is the **uniform distribution** on $(0, 1)$.

### Example 10 — Triangular Distribution {#ex10}

**Problem:** $f(x_1, x_2) = 1$ for $0 < x_1 < 1$, $0 < x_2 < 1$ (0 elsewhere). Find the marginal density of $Y = X_1 + X_2$.

**Solution:**

Let $Y = X_1 + X_2$ and $Z = X_2$. Then $x_1 = y - z$, $x_2 = z$, $|J| = 1$.

The region maps to $z < y < z+1$, $0 < z < 1$, i.e., $z < y < z+1$.

Integrating out $z$:

$$h(y) = \begin{cases} \int_0^y dz = y, & 0 \leq y < 1 \\[4pt] \int_{y-1}^{1} dz = 2 - y, & 1 \leq y \leq 2 \\[4pt] 0, & \text{elsewhere} \end{cases}$$

This is the **triangular distribution**.

### Example 11 — Three Variables {#ex11}

**Problem:** $f(x_1, x_2, x_3) = e^{-(x_1+x_2+x_3)}$ for $x_i > 0$ (0 elsewhere). Find the marginal density of $Y_1 = X_1 + X_2 + X_3$.

**Solution:**

Let $Y_2 = X_2$ and $Y_3 = X_3$. Then $x_1 = y_1 - y_2 - y_3$, $x_2 = y_2$, $x_3 = y_3$. The $3\times 3$ Jacobian equals 1.

$$g(y_1, y_2, y_3) = e^{-y_1}, \quad y_1 > y_2 + y_3,\ y_2, y_3 > 0$$

Integrating out $y_2$ and $y_3$:

$$h(y_1) = \int_0^{y_1}\int_0^{y_1 - y_3} e^{-y_1}\, dy_2\, dy_3 = \frac{1}{2}y_1^2 e^{-y_1}$$

$$\boxed{h(y_1) = \begin{cases} \dfrac{1}{2}y_1^2 e^{-y_1}, & y_1 > 0 \\ 0, & \text{elsewhere} \end{cases}}$$

This is a **Gamma distribution** with $\alpha = 3$ and $\beta = 1$.

---

## Exercises — Week 5 {#ex-week5}

1. If $X$ has a hypergeometric distribution with $m=3$, $N=6$, $n=2$, find the p.d. of $Z = (X-1)^2$.
   $$h(0) = \frac{3}{5}, \quad h(1) = \frac{2}{5}$$

2. If $X$ has a binomial distribution with $n=3$, $p=\frac{1}{3}$, find the p.d. of:
   - (a) $Y = \dfrac{X}{X+1}$: $g(0) = \frac{8}{27}$, $g(\frac{1}{2}) = \frac{12}{27}$, $g(\frac{2}{3}) = \frac{6}{27}$, $g(\frac{3}{4}) = \frac{1}{27}$
   - (b) $U = (X-1)^4$: $g(0) = \frac{12}{27}$, $g(1) = \frac{14}{27}$, $g(16) = \frac{1}{27}$

3. If $X = \ln(Y)$ has a normal distribution with mean $\mu$ and std dev $\sigma$, find the p.d.f. of $Y$ (the **log-normal distribution**):
$$g(y) = \frac{1}{\sqrt{2\pi}\,\sigma y} \exp\!\left(-\frac{(\ln y - \mu)^2}{2\sigma^2}\right), \quad y > 0$$

4. If $f(x) = kx^3(1-x^2)^6$ for $0 < x < 1$ (0 elsewhere), find the p.d.f. of $Y = X^2$. Show it is a Beta distribution and find $k$.

   The distribution is Beta with $\alpha = 4$, $\beta = 2$; $k = 320$.

5. If $X$ has a uniform density with $\alpha=0$, $\beta=1$, show that $Y = -2\ln X$ has a **gamma distribution**. Find its parameters.

6. If $X$ has a uniform density with $\alpha=0$, $\beta=1$, show that $Y = X^{-1/\alpha}$ (for $\alpha > 0$) has the **Pareto distribution**.

7. Consider $X \sim \text{Uniform}(-1, 3)$:
   - (a) Find the p.d.f. of $Y = |X|$:
     $g(y) = \frac{1}{4}$ for $0 \leq y < 1$; $g(y) = \frac{1}{4}$ for $1 \leq y < 3$
   - (b) Find the p.d.f. of $Z = (X-Y)^2 = (X - |X|)^2$.

8. If $f(x_1, x_2) = \dfrac{x_1 x_2}{36}$ for $x_1, x_2 = 1, 2, 3$, find the p.d. of:
   - (a) $X_1 X_2$
   - (b) $X_1 / X_2$

9. With the same joint distribution as in Exercise 8, find:
   - (a) The joint distribution of $Y_1 = X_1 + X_2$ and $Y_2 = X_1 - X_2$
   - (b) The marginal distribution of $Y_1$:
     $g(2) = \frac{1}{36}$, $g(3) = \frac{4}{36}$, $g(4) = \frac{10}{36}$, $g(5) = \frac{12}{36}$, $g(6) = \frac{9}{36}$

10. If $X_1$, $X_2$, $X_3$ have the multinomial distribution with $n=2$, $\theta_1=\frac{1}{4}$, $\theta_2=\frac{1}{2}$, $\theta_3=\frac{5}{12}$, find the joint p.d. of $Y_1 = X_1 + X_2$, $Y_2 = X_1 - X_2$, $Y_3 = X_3$.

11. If $X_1$ and $X_2$ are independent binomial with parameters $(n_1, \theta)$ and $(n_2, \theta)$, show that $Y = X_1 + X_2$ is binomial with parameters $(n_1 + n_2, \theta)$.

12. If $X$ and $Y$ are independent standard normal, show that $Z = X + Y$ is normally distributed. Find the mean and variance.
    $\mu = 0$, $\sigma^2 = 2$

13. Given $f(x, y) = 12y(1-x-y)$ for $0 < x < 1$, $0 < y < 1-x$ (0 elsewhere):
    - (a) Find the joint density of $Z = XY$ and $U = Y$
    - (b) Find the marginal density of $Z$: $h(z) = 6z - 6z^2 - 12z^2$ for $0 < z < 1$

14. Two independent Cauchy random variables $X_1$ and $X_2$ with density $f(x) = \dfrac{1}{\pi(1+x^2)}$:
    - (a) Find the joint density of $Y_1 = X_1 + X_2$ and $Y_2 = X_1 - X_2$
    - (b) Find the marginal density of $Y_1$: $g(y_1) = \dfrac{1}{\pi(4 + y_1^2)} \cdot 4$

15. Given $f(x,y) = 24xy$ for $0 < x < 1$, $0 < y < 1$, $x + y < 1$, find the joint density of $Z = X+Y$ and $W = X$.

16. Let $X$ and $Y$ be independent gamma random variables. Find the joint density of $U = \dfrac{X}{X+Y}$ and $V = X+Y$, and identify the marginal density of $U$.

17. The Maxwell–Boltzmann velocity distribution is $f(v) = kv^2 e^{-\beta v^2}$ for $v > 0$. Show that kinetic energy $E = \frac{1}{2}mV^2$ has a **gamma distribution**.

---

# Moment-Generating Function Technique {#week6}

## Overview {#mgf-overview}

Moment-generating functions provide an elegant tool for finding the distribution of **linear combinations of independent random variables**.

The method is based on the uniqueness of the MGF: if $M_Y(t) = M_Z(t)$ for all $t$ in some interval, then $Y$ and $Z$ have the same distribution.

## Theorem 3 — MGF of a Sum {#theorem3}

> **Theorem 3:** If $X_1, X_2, \ldots, X_n$ are **independent** random variables and $Y = X_1 + X_2 + \cdots + X_n$, then:
>
> $$M_Y(t) = \prod_{i=1}^{n} M_{X_i}(t)$$

**Proof:**

$$M_Y(t) = E\!\left[e^{tY}\right] = E\!\left[e^{t(X_1 + \cdots + X_n)}\right] = E\!\left[e^{tX_1}\right] \cdots E\!\left[e^{tX_n}\right] = \prod_{i=1}^{n} M_{X_i}(t)$$

(using independence). $\square$

## Example 12 — Sum of Poisson Variables {#ex12}

**Problem:** $X_1, \ldots, X_n$ are independent Poisson with parameters $\lambda_1, \ldots, \lambda_n$. Find the distribution of $Y = X_1 + \cdots + X_n$.

**Solution:**

The MGF of a Poisson$(\lambda_i)$ is $M_{X_i}(t) = e^{\lambda_i(e^t - 1)}$. By Theorem 3:

$$M_Y(t) = \prod_{i=1}^n e^{\lambda_i(e^t-1)} = e^{(\lambda_1 + \cdots + \lambda_n)(e^t - 1)}$$

> **Conclusion:** $Y \sim \text{Poisson}(\lambda_1 + \lambda_2 + \cdots + \lambda_n)$.

## Example 13 — Sum of Exponential Variables {#ex13}

**Problem:** $X_1, \ldots, X_n$ are i.i.d. Exponential$(\lambda)$ (equivalently Gamma$(1, \lambda)$). Find the distribution of $Y = \sum X_i$.

**Solution:**

The MGF of Gamma$(\alpha, \beta)$ is $M_X(t) = \left(1 - \frac{t}{\beta}\right)^{-\alpha}$. For Exponential$(\lambda)$:

$$M_{X_i}(t) = \left(1 - \frac{t}{\lambda}\right)^{-1}$$

$$M_Y(t) = \left(1 - \frac{t}{\lambda}\right)^{-n}$$

> **Conclusion:** $Y \sim \text{Gamma}(n, \lambda)$ with p.d.f.:
>
> $$g(y) = \frac{\lambda^n}{\Gamma(n)} y^{n-1} e^{-\lambda y}, \quad y > 0$$

## Applied Example — Lake Victoria Fish {#ex-fish}

**Setup:** The number of fish caught per hour at Lake Victoria follows $\text{Poisson}(\lambda = 1.6)$.

### (a) Four fish in two hours {-}

Let $Y = X_1 + X_2 \sim \text{Poisson}(3.2)$:

$$P(Y = 4) = \frac{e^{-3.2}(3.2)^4}{4!} = \frac{0.04076 \times 104.86}{24} \approx \mathbf{0.1781}$$

### (b) At least two fish in three hours {-}

Let $Y = X_1 + X_2 + X_3 \sim \text{Poisson}(4.8)$:

$$P(Y \geq 2) = 1 - P(Y = 0) - P(Y = 1)$$

$$P(Y = 0) = e^{-4.8} \approx 0.00823, \quad P(Y = 1) = 4.8e^{-4.8} \approx 0.03950$$

$$P(Y \geq 2) = 1 - 0.04773 \approx \mathbf{0.9523}$$

### (c) At most three fish in four hours {-}

Let $Y = X_1 + X_2 + X_3 + X_4 \sim \text{Poisson}(6.4)$:

$$P(Y \leq 3) = \sum_{y=0}^{3} \frac{e^{-6.4}(6.4)^y}{y!}$$

$$= 0.00166 + 0.01063 + 0.03403 + 0.07253 \approx \mathbf{0.1189}$$

---

## Chebyshev's Theorem {#chebyshev}

### Statement {#cheb-statement}

> **Chebyshev's Theorem:** If $\mu$ and $\sigma$ are the mean and standard deviation of a random variable $X$, then for any $k > 0$:
>
> $$\Pr(|X - \mu| < k\sigma) \geq 1 - \frac{1}{k^2}$$
>
> Equivalently: $\Pr(\mu - k\sigma < X < \mu + k\sigma) \geq 1 - \dfrac{1}{k^2}$.

This provides a **universal lower bound** on the probability that $X$ lies within $k$ standard deviations of its mean, valid for *any* distribution.

### Proof {#cheb-proof}

By definition:

$$\sigma^2 = E[(X - \mu)^2] = \int_{-\infty}^{\infty}(x-\mu)^2 f(x)\,dx$$

Split the integral into three regions:

$$\sigma^2 \geq \int_{-\infty}^{\mu - k\sigma}(x-\mu)^2 f(x)\,dx + \int_{\mu + k\sigma}^{\infty}(x-\mu)^2 f(x)\,dx$$

Since $(x - \mu)^2 \geq k^2\sigma^2$ in the outer regions:

$$\sigma^2 \geq k^2\sigma^2 \cdot \Pr(|X - \mu| \geq k\sigma)$$

$$\Pr(|X - \mu| \geq k\sigma) \leq \frac{1}{k^2}$$

$$\Pr(|X - \mu| < k\sigma) \geq 1 - \frac{1}{k^2} \quad \square$$

### Example 1 — Beta Distribution {#cheb-ex1}

**Problem:** $X$ has p.d.f. $f(x) = 630x^4(1-x)^4$ for $0 < x < 1$. Compare the exact probability within two standard deviations of the mean to Chebyshev's bound.

**Solution:**

Comparing to the Beta$(\alpha, \beta)$ density, we identify $\alpha = 5$, $\beta = 5$.

$$\mu = E(X) = \frac{\alpha}{\alpha + \beta} = \frac{5}{10} = 0.5$$

$$\sigma^2 = \text{Var}(X) = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)} = \frac{25}{100 \times 11} = \frac{1}{44}$$

$$\sigma = \frac{1}{\sqrt{44}} \approx 0.1508$$

The interval within 2 standard deviations: $[0.5 - 2(0.1508),\ 0.5 + 2(0.1508)] = [0.20, 0.80]$.

**Exact probability:**

$$P(0.20 < X < 0.80) = 630\int_{0.20}^{0.80} x^4(1-x)^4\, dx \approx \mathbf{0.96}$$

**Chebyshev bound** (with $k = 2$):

$$1 - \frac{1}{k^2} = 1 - \frac{1}{4} = 0.75$$

> The exact probability $0.96$ is much stronger than Chebyshev's lower bound $0.75$.

### Example 2 — Uniform Distribution {#cheb-ex2}

**Problem:** $X \sim \text{Uniform}(-3, 3)$. With $k = \frac{3}{2}$, find:

- (a) The exact probability
- (b) The Chebyshev bound

**Solution:**

$$\mu = \frac{-3 + 3}{2} = 0, \qquad \sigma^2 = \frac{(3 - (-3))^2}{12} = \frac{36}{12} = 3, \qquad \sigma = \sqrt{3}$$

With $k = \frac{3}{2}$: $k\sigma = \frac{3}{2}\sqrt{3}$

**Exact probability:**

$$P\!\left(|X| < \frac{3\sqrt{3}}{2}\right) = P\!\left(-\frac{3\sqrt{3}}{2} < X < \frac{3\sqrt{3}}{2}\right) = \frac{1}{6} \cdot 3\sqrt{3} \approx 0.866$$

Actually, since $\frac{3\sqrt{3}}{2} \approx 2.598 < 3$:

$$P = \frac{2 \cdot 2.598}{6} = \frac{5.196}{6} \approx \mathbf{0.866}$$

**Chebyshev bound:**

$$1 - \frac{1}{(3/2)^2} = 1 - \frac{4}{9} = \frac{5}{9} \approx 0.556$$

---

## Exercises — Week 6 {#ex-week6}

1.  **MGF Technique:** If $X_1$ and $X_2$ are independent binomial with parameters $(n_1, \theta)$ and $(n_2, \theta)$, show using MGFs that $Y = X_1 + X_2 \sim \text{Binomial}(n_1 + n_2, \theta)$.

2. If $n$ independent random variables each have Gamma$(\alpha, \beta)$ distributions, find the MGF of their sum and identify the distribution.
   **Answer:** Gamma$(n\alpha, \beta)$.

3. If $X_i \sim N(\mu_i, \sigma_i^2)$ independently, find the MGF of $Y = \sum X_i$ and identify the distribution.
   $$Y \sim N\!\left(\sum_{i=1}^n \mu_i,\ \sum_{i=1}^n \sigma_i^2\right)$$

4. Prove the generalization of Theorem 3: if $Y = a_1 X_1 + \cdots + a_n X_n$ with $X_i$ independent, then $M_Y(t) = \prod_{i=1}^n M_{X_i}(a_i t)$.

5. A lawyer receives calls on an unlisted number at rate $2.1$/half-hour and on a listed number at rate $10.9$/half-hour (independent Poisson). Find the probabilities that in half an hour she receives:
   - (a) 14 calls: **0.1021**
   - (b) At most 6 calls: **0.0259**

6. (Repeat of fish problem — see Section \@ref(ex-fish) above.)

7. If the time a doctor spends per patient is Exponential$(\lambda = \frac{1}{9})$ (minutes), find the probability the doctor spends at least 20 minutes with:
   - (a) One patient: $P(X_1 \geq 20) = e^{-20/9} \approx \mathbf{0.1084}$
   - (b) Two patients: $P(Y \geq 20)$ where $Y \sim \text{Gamma}(2, 1/9)$: **0.3492**
   - (c) Three patients: $P(Y \geq 20)$ where $Y \sim \text{Gamma}(3, 1/9)$: **0.6168**

8. **Chebyshev:** Find the smallest $k$ such that $P(|X - \mu| < k\sigma) \geq$:
   - (a) 0.95: $k = \sqrt{20} \approx 4.47$
   - (b) 0.99: $k = \sqrt{100} = 10$

9. The thiamine content in a bread slice has $\mu = 0.260$ mg, $\sigma = 0.005$ mg. By Chebyshev's theorem, between what values must the thiamine content lie for:
   - (a) At least $\frac{35}{36}$ of all slices: $[0.230,\ 0.290]$ mg
   - (b) At least $\frac{143}{144}$ of all slices: $[0.200,\ 0.320]$ mg

10. If $E(X) = 3$ and $E(X^2) = \frac{13}{2}$, use Chebyshev's theorem to find a lower bound for $P(2 < X < 8)$.
    **Answer:** 0.84

---

*End of STA227 Weeks 4–6 Notes*

# Week 7: Sampling Distribution {#week7}

## Introduction {#introduction}

Statistics concerns itself mainly with conclusions and predictions resulting from chance outcomes that occur in carefully planned experiments or investigations.

In the **finite case**, these chance outcomes constitute a subset, or **sample**, of measurements or observations from a larger set of values called the **population**. In the **continuous case** they are usually identically distributed random variables, whose distribution we refer to as the **population distribution**, or the *infinite population* sampled.

> *Not all samples lend themselves to valid generalizations about the populations from which they came. Most methods of inference are based on the assumption that we are dealing with* ***random samples***.

### Definition 1 – Random Sample

If $X_1, X_2, \ldots, X_n$ are **independent and identically distributed** random variables, we say they constitute a **random sample** from the infinite population given by their common distribution.

If $f(x_1, x_2, \ldots, x_n)$ is the value of the joint distribution, we can write:

$$f(x_1, x_2, \ldots, x_n) = \prod_{i=1}^{n} f(x_i)$$

where $f(x_i)$ is the value of the population distribution at $x_i$.

Statistical inferences are usually based on **statistics** — random variables that are functions of $X_1, X_2, \ldots, X_n$.

### Definition 2 – Sample Mean and Sample Variance

If $X_1, X_2, \ldots, X_n$ constitute a random sample, then:

$$\bar{X} = \frac{1}{n}\sum_{i=1}^{n} X_i \quad \text{(sample mean)}$$

$$S^2 = \frac{\sum_{i=1}^{n}(X_i - \bar{X})^2}{n - 1} \quad \text{(sample variance)}$$

For observed sample data, we calculate $\bar{x}$ and $s^2$; these are values of the corresponding random variables $\bar{X}$ and $S^2$.

---

## The Distribution of the Mean {#distribution-mean}

Since statistics are random variables, their values vary from sample to sample. Their distributions are called **sampling distributions**.

### Theorem 1 – Mean and Variance of $\bar{X}$

If $X_1, X_2, \ldots, X_n$ constitute a random sample from an infinite population with mean $\mu$ and variance $\sigma^2$, then:

$$E(\bar{X}) = \mu \qquad \text{and} \qquad \text{Var}(\bar{X}) = \frac{\sigma^2}{n}$$

**Proof:**

$$E(\bar{X}) = E\!\left(\frac{1}{n}\sum_{i=1}^n X_i\right) = \frac{1}{n}\sum_{i=1}^n E(X_i) = \frac{1}{n} \cdot n\mu = \mu$$

$$\text{Var}(\bar{X}) = \text{Var}\!\left(\frac{1}{n}\sum_{i=1}^n X_i\right) = \frac{1}{n^2}\sum_{i=1}^n \text{Var}(X_i) = \frac{1}{n^2} \cdot n\sigma^2 = \frac{\sigma^2}{n}$$

> The **standard error of the mean** is $\sigma_{\bar{X}} = \dfrac{\sigma}{\sqrt{n}}$. As $n$ increases, $\sigma_{\bar{X}}$ decreases — larger samples yield $\bar{X}$ values closer to $\mu$.

---

### Theorem 2 – Chebyshev's Bound for $\bar{X}$

For any positive constant $c$, the probability that $\bar{X}$ falls between $\mu - c$ and $\mu + c$ is at least:

$$1 - \frac{\sigma^2}{nc^2}$$

As $n \to \infty$, this probability approaches **1**.

**Proof:**

From Chebyshev's theorem, for any random variable with mean $\mu$ and standard deviation $\sigma$, and any $k > 0$:

$$P(|X - \mu| < k\sigma) \geq 1 - \frac{1}{k^2}$$

Applying this to $\bar{X}$ (which has standard deviation $\sigma/\sqrt{n}$), set $k\sigma_{\bar{X}} = c$, so $k = \dfrac{c\sqrt{n}}{\sigma}$:

$$P(|\bar{X} - \mu| < c) \geq 1 - \frac{\sigma^2}{nc^2}$$

This result is known as the **Law of Large Numbers**. $\blacksquare$

---

## Central Limit Theorem {#clt}

### Theorem 3 – Central Limit Theorem (CLT)

If $X_1, X_2, \ldots, X_n$ constitute a random sample from an infinite population with mean $\mu$, variance $\sigma^2$, and moment-generating function $M_X(t)$, then the **limiting distribution** of:

$$Z = \frac{\bar{X} - \mu}{\sigma / \sqrt{n}}$$

as $n \to \infty$ is the **standard normal distribution**.

> **Practical rule:** The CLT approximation is used when $n \geq 30$, regardless of the shape of the population.

> **Important note:** The CLT does *not* say the distribution of $\bar{X}$ becomes normal (since $\text{Var}(\bar{X}) \to 0$). It justifies approximating $\bar{X}$ with a normal having mean $\mu$ and variance $\sigma^2/n$ when $n$ is large.

---

### Example 1 – Soft Drink Vending Machine

A vending machine dispenses drinks with mean $\mu = 200\,\text{ml}$ and standard deviation $\sigma = 15\,\text{ml}$. Find $P(\bar{X} \geq 204)$ for $n = 36$.

**Solution:**

$$\sigma_{\bar{X}} = \frac{15}{\sqrt{36}} = 2.5$$

$$P(\bar{X} \geq 204) = P\!\left(Z \geq \frac{204 - 200}{2.5}\right) = P(Z \geq 1.6)$$

$$= 1 - P(Z \leq 1.6) = 1 - 0.9452 = \boxed{0.0548}$$

---

### Example 2 – Gamma Population

A random sample of size $n = 72$ is taken from:

$$f(x) = \frac{1}{16}x\,e^{-x/4}, \quad x > 0$$

Use the CLT to find $P(\bar{X} > 9)$.

**Solution:**

Identifying this as a Gamma distribution with $\alpha = 2$, $\beta = 4$:

$$E(X) = \alpha\beta = 8, \qquad \text{Var}(X) = \alpha\beta^2 = 32$$

$$P(\bar{X} > 9) = P\!\left(Z > \frac{9 - 8}{\sqrt{32/72}}\right) = P(Z > 1.5) = 1 - 0.9332 = \boxed{0.0668}$$

---

### Theorem 4 – Exact Distribution for Normal Populations

If $\bar{X}$ is the mean of a random sample of size $n$ from a **normal** population with mean $\mu$ and variance $\sigma^2$, then the **exact** sampling distribution is:

$$\bar{X} \sim N\!\left(\mu,\; \frac{\sigma^2}{n}\right)$$

(This holds for *any* $n$, without the CLT approximation.)

---

## Exercises – Week 7 {#exercises-week7}

1. A random sample of size $n = 100$ is taken from a population with $\mu = 75$ and $\sigma^2 = 256$.
   - **(a)** Using Chebyshev's theorem, find $P(67 < \bar{X} < 83)$. **[Answer: 0.96]**
   - **(b)** Using the CLT, find $P(67 < \bar{X} < 83)$. **[Answer: 0.9999994]**

2. A random sample of size $n = 81$ is taken from a population with $\mu = 128$ and $\sigma = 6.3$. Find $P(\bar{X} \notin (126.6,\; 129.4))$ using:
   - **(a)** Chebyshev's theorem. **[At most 0.25]**
   - **(b)** The CLT. **[0.0456]**

3. A random sample of size 64 from a normal population with $\mu = 51.4$, $\sigma = 6.8$. Find:
   - **(a)** $P(\bar{X} > 52.9)$ **[0.0388]**
   - **(b)** $P(50.5 < \bar{X} < 52.3)$ **[0.7108]**
   - **(c)** $P(\bar{X} < 50.6)$ **[0.1736]**

4. $n = 225$ from an exponential population with $\theta = 4$. Find $P(\bar{X} > 4.5)$ using the CLT.

5. $n = 200$ from a uniform population with $\alpha = 24$, $\beta = 48$. Find $P(\bar{X} < 35)$. **[0.0207]**

6. $n = 100$ from a normal population with $\sigma = 25$. Find $P(|\bar{X} - \mu| \geq 3)$. **[0.2302]**

7. Let $\bar{X}$ be the mean of a random sample of size 100 from a distribution with $\sigma^2 = 50$. Find approximately $P(49 < \bar{X} < 51)$.

8. Let $f(x) = \frac{1}{x^2}$ for $x \geq 1$. For $n = 72$, find approximately the probability that more than 50 items are less than 3. **[0.267]**

9. $\bar{X}$ is the mean of a random sample of size 128 from a Gamma$(\alpha=2, \beta=4)$ distribution. Find approximately $P(7 < \bar{X} < 9)$. **[0.954]**

10. Find the approximate probability that the mean of a sample of size 15 from $f(x) = 3x^2$, $0 < x < 1$, lies between $\frac{3}{5}$ and $\frac{4}{5}$. **[0.840]**

---

# Week 8: The Chi-Square and t-Distributions {#week8}

## The Chi-Square Distribution {#chi-square}

A random variable $X$ has the **Chi-square distribution** with $r$ degrees of freedom if its pdf is:

$$f(x) = \frac{1}{2^{r/2}\,\Gamma(r/2)}\,x^{r/2 - 1}\,e^{-x/2}, \quad x > 0$$

We write $X \sim \chi^2_r$ (or $X$ is $\chi^2_r$).

---

### MGF of the Chi-Square Distribution

$$M_X(t) = (1 - 2t)^{-r/2}, \quad t < \frac{1}{2}$$

**Proof sketch:** Substituting $y = x(1-2t)/2$ into the integral and using the Gamma function identity $\int_0^\infty y^{r/2-1}e^{-y}\,dy = \Gamma(r/2)$ gives the result above.

---

### Mean and Variance of the Chi-Square Distribution

Differentiating $M_X(t) = (1-2t)^{-r/2}$:

$$E(X) = r \qquad \text{and} \qquad \text{Var}(X) = 2r$$

---

### Theorem 5 – Square of Standard Normal

If $Z \sim N(0,1)$, then $Z^2 \sim \chi^2_1$.

**Proof:** The MGF of $Z^2$ is derived as $(1-2t)^{-1/2}$, which is the MGF of $\chi^2_1$. $\blacksquare$

---

### Theorem 6 – Sum of Squared Standard Normals

If $X_1, X_2, \ldots, X_n$ are independent $N(0,1)$ random variables, then:

$$Y = \sum_{i=1}^{n} X_i^2 \sim \chi^2_n$$

**Proof:** By Theorem 5 each $X_i^2 \sim \chi^2_1$, and since they are independent:

$$M_Y(t) = \prod_{i=1}^n (1-2t)^{-1/2} = (1-2t)^{-n/2}$$

which is the MGF of $\chi^2_n$. $\blacksquare$

---

### Theorem 7 – Reproductive Property

If $X_1, \ldots, X_n$ are independent with $X_i \sim \chi^2_{r_i}$, then:

$$Y = \sum_{i=1}^n X_i \sim \chi^2_{r_1 + r_2 + \cdots + r_n}$$

---

### Theorem 8 – Difference of Chi-Square Variables

If $X_1$ and $X_2$ are independent, $X_1 \sim \chi^2_{r_1}$, and $X_1 + X_2 \sim \chi^2_{r_1+r}$, then $X_2 \sim \chi^2_r$.

---

### Theorem 9 – Sample Mean and Variance from a Normal Population

If $\bar{X}$ and $S^2$ are the mean and variance of a random sample of size $n$ from $N(\mu, \sigma^2)$, then:

**(a)** $\bar{X}$ and $S^2$ are **independent**.

**(b)** $\dfrac{(n-1)S^2}{\sigma^2} \sim \chi^2_{n-1}$

**Proof of (b):** Using the identity:

$$\sum_{i=1}^n \frac{(X_i - \mu)^2}{\sigma^2} = \frac{(n-1)S^2}{\sigma^2} + \left(\frac{\bar{X} - \mu}{\sigma/\sqrt{n}}\right)^2$$

The left side is $\chi^2_n$ (by Theorem 6). The second term on the right is $\chi^2_1$ (by Theorems 4 and 5). By Theorem 8, $\dfrac{(n-1)S^2}{\sigma^2} \sim \chi^2_{n-1}$. $\blacksquare$

---

### Example 2 – Semiconductor Thickness

A manufacturing process is "in control" if $\sigma \leq 0.60$ thousandths of an inch. For $n = 20$, the process is declared "out of control" if $\dfrac{(n-1)S^2}{\sigma^2} \geq \chi^2_{0.01,\,19} = 36.191$.

With $S = 0.84$, $\sigma = 0.60$:

$$\frac{(n-1)S^2}{\sigma^2} = \frac{19 \times (0.84)^2}{(0.60)^2} = 37.24 > 36.191$$

**Conclusion:** Reject $H_0$; the process is **out of control**.

---

## The t-Distribution {#t-distribution}

Although $Z = \dfrac{\bar{X} - \mu}{\sigma/\sqrt{n}} \sim N(0,1)$ is elegant, in practice $\sigma$ is usually **unknown** and must be replaced by the sample standard deviation $S$.

### Theorem 10 – Definition of the t-Distribution

If $Y \sim \chi^2_r$ and $Z \sim N(0,1)$ are independent, then:

$$T = \frac{Z}{\sqrt{Y/r}}$$

has the **t-distribution** with $r$ degrees of freedom, with pdf:

$$f(t) = \frac{\Gamma\!\left(\frac{r+1}{2}\right)}{\sqrt{r\pi}\;\Gamma\!\left(\frac{r}{2}\right)} \left(1 + \frac{t^2}{r}\right)^{-(r+1)/2}, \quad -\infty < t < \infty$$

> Originally introduced by **W.S. Gosset** under the pen-name **"Student"** (his employer, a brewery, did not permit employee publications). Hence also known as **Student's t-distribution**.

---

### Theorem 11 – t-Statistic for a Normal Sample

If $\bar{X}$ and $S^2$ are the mean and variance of a random sample of size $n$ from $N(\mu, \sigma^2)$, then:

$$T = \frac{\bar{X} - \mu}{S/\sqrt{n}} \sim t_{n-1}$$

**Proof:** By Theorem 9, set $Z = \dfrac{\bar{X}-\mu}{\sigma/\sqrt{n}} \sim N(0,1)$ and $Y = \dfrac{(n-1)S^2}{\sigma^2} \sim \chi^2_{n-1}$, which are independent. Substituting into Theorem 10:

$$T = \frac{Z}{\sqrt{Y/(n-1)}} = \frac{(\bar{X}-\mu)/(\sigma/\sqrt{n})}{\sqrt{(n-1)S^2/[\sigma^2(n-1)]}} = \frac{\bar{X}-\mu}{S/\sqrt{n}} \sim t_{n-1} \quad \blacksquare$$

---

### Example 3 – Gasoline Consumption

In 16 one-hour test runs, an engine averaged $\bar{x} = 16.4$ gallons with $s = 2.1$ gallons. Test the claim that $\mu = 12.0$ gallons per hour.

**Solution:**

$$t = \frac{\bar{x} - \mu_0}{s/\sqrt{n}} = \frac{16.4 - 12.0}{2.1/\sqrt{16}} = \frac{4.4}{0.525} = 8.38$$

From t-tables: $t_{0.005,\,15} = 2.947$.

Since $8.38 > 2.947$, we **reject the claim**. The true average consumption exceeds 12.0 gallons per hour.

---

# Week 9: The F-Distribution {#week9}

## The F-Distribution {#f-distribution}

The **F-distribution** is named after **Sir Ronald A. Fisher**, one of the most prominent statisticians of the 20th century. It was originally studied as the sampling distribution of the ratio of two independent chi-square random variables, each divided by its degrees of freedom.

### Theorem 12 – Definition of the F-Distribution

If $U \sim \chi^2_{n_1}$ and $V \sim \chi^2_{n_2}$ are **independent**, then:

$$F = \frac{U/n_1}{V/n_2}$$

has an **F-distribution** with $n_1$ and $n_2$ degrees of freedom, with pdf:

$$g(f) = \frac{\Gamma\!\left(\frac{n_1+n_2}{2}\right)}{\Gamma\!\left(\frac{n_1}{2}\right)\Gamma\!\left(\frac{n_2}{2}\right)} \left(\frac{n_1}{n_2}\right)^{n_1/2} \frac{f^{n_1/2 - 1}}{\left(1 + \dfrac{n_1}{n_2}f\right)^{(n_1+n_2)/2}}, \quad f > 0$$

**Proof:** Apply the change-of-variable $f = \dfrac{un_2}{vn_1}$ to the joint density of $U$ and $V$, then integrate out $v$ using the substitution $w = v\!\left(1 + \dfrac{n_1}{n_2}f\right)/2$. $\blacksquare$

---

## Mean and Variance of the F-Distribution {#f-mean-variance}

Let $X \sim F(m, n)$. Then:

$$E(X) = \frac{n}{n-2}, \quad n > 2$$

$$\text{Var}(X) = \frac{2n^2(m+n-2)}{m(n-2)^2(n-4)}, \quad n > 4$$

**Proof sketch:**

Write $X = \dfrac{U/m}{V/n}$ where $U \sim \chi^2_m$ and $V \sim \chi^2_n$ are independent.

$$E(X) = \frac{n}{m} E(U) \cdot E\!\left(\frac{1}{V}\right) = \frac{n}{m} \cdot m \cdot \frac{1}{n-2} = \frac{n}{n-2}$$

since $E(1/V) = 1/(n-2)$ for $V \sim \chi^2_n$ (obtained by substitution in the integral). Similarly, $E(1/V^2) = 1/[(n-2)(n-4)]$ for $n > 4$, which yields the variance formula. $\blacksquare$

---

## Applications of the F-Distribution {#f-applications}

The F-distribution arises naturally when **comparing variances** $\sigma_1^2$ and $\sigma_2^2$ of two normal populations.

### Theorem 13 – Ratio of Sample Variances

If $S_1^2$ and $S_2^2$ are the variances of **independent** random samples of sizes $n_1$ and $n_2$ from $N(\mu_1, \sigma_1^2)$ and $N(\mu_2, \sigma_2^2)$ respectively, then:

$$F = \frac{S_1^2 / \sigma_1^2}{S_2^2 / \sigma_2^2} \sim F(n_1-1,\; n_2-1)$$

**Proof:** By Theorem 9, $\dfrac{(n_1-1)S_1^2}{\sigma_1^2} \sim \chi^2_{n_1-1}$ and $\dfrac{(n_2-1)S_2^2}{\sigma_2^2} \sim \chi^2_{n_2-1}$, independently. Substituting into Theorem 12 gives the result. $\blacksquare$

> The F-table gives critical values $f_\alpha(n_1, n_2)$ such that $P(F > f_\alpha) = \alpha$ for specified degrees of freedom.

---

## Exercises – Weeks 8 & 9 {#exercises-weeks89}

1. Use Theorem 9 to show that for random samples of size $n$ from $N(\mu, \sigma^2)$, the sampling distribution of $S^2$ has mean $\sigma^2$ and variance $\dfrac{2\sigma^4}{n-1}$.

2. Show that if $X_1, X_2, \ldots, X_n$ are independent $\chi^2_1$ and $Y_n = X_1 + X_2 + \cdots + X_n$, then the limiting distribution of:
$$Z_n = \frac{Y_n - n}{\sqrt{2n}}$$
is $N(0,1)$ as $n \to \infty$.

3. Using Exercise 2, show that if $X \sim \chi^2_n$ with large $n$, then:
$$\frac{X - n}{\sqrt{2n}} \approx N(0,1)$$

4. Use Exercise 3 to find the approximate probability that a $\chi^2_{50}$ random variable exceeds **68.0**.

5. Show that for $n > 2$, the variance of the $t$-distribution with $n$ degrees of freedom is $\dfrac{n}{n-2}$. *(Hint: substitute $t = \sqrt{\dfrac{n}{1-u^2}} \cdot \text{sign}(t)$)*

6. Verify that if $T \sim t_n$, then $T^2 \sim F(1, n)$.

7. Verify that if $X \sim F(n_1, n_2)$ and $n_2 \to \infty$, the distribution of $Y = n_1 X$ approaches $\chi^2_{n_1}$.

8. If $X \sim F(n_1, n_2)$, show that $Y = \dfrac{1}{X} \sim F(n_2, n_1)$.

9. Verify that if $Y$ has a Beta distribution with $\alpha = \dfrac{n_1}{2}$ and $\beta = \dfrac{n_2}{2}$, then:
$$X = \frac{n_2 Y}{n_1(1-Y)} \sim F(n_1, n_2)$$

---

*End of STA227 Weeks 7–9 Notes*

# Week 10: Order Statistics {#week10}

## Introduction {#intro-order}

Consider a random sample of size $n$ from an infinite population with a **continuous density**, and suppose we arrange the values $X_1, X_2, \ldots, X_n$ according to size.

- The **smallest** value is the random variable $Y_1$
- The **second smallest** is $Y_2$
- $\vdots$
- The **largest** is $Y_n$

These $Y$'s are called **order statistics**. In particular, $Y_r$ is called the $r^{\text{th}}$ order statistic.

> *We restrict to infinite populations with continuous densities so there is zero probability that any two sample values are equal.*

### Case $n = 2$

$$y_1 = x_1,\; y_2 = x_2 \quad \text{when } x_1 < x_2$$
$$y_1 = x_2,\; y_2 = x_1 \quad \text{when } x_2 < x_1$$

### Case $n = 3$

| Condition | $y_1$ | $y_2$ | $y_3$ |
|:---|:---:|:---:|:---:|
| $x_1 < x_2 < x_3$ | $x_1$ | $x_2$ | $x_3$ |
| $x_1 < x_3 < x_2$ | $x_1$ | $x_3$ | $x_2$ |
| $x_2 < x_1 < x_3$ | $x_2$ | $x_1$ | $x_3$ |

---

## Theorem 14 – pdf of the $r^{\text{th}}$ Order Statistic {#theorem14}

For random samples of size $n$ from an infinite population with pdf $f(x)$, the probability density of the $r^{\text{th}}$ order statistic $Y_r$ is:

$$g_r(y_r) = \frac{n!}{(r-1)!\,(n-r)!} \left[\int_{-\infty}^{y_r} f(x)\,dx\right]^{r-1} f(y_r) \left[\int_{y_r}^{\infty} f(x)\,dx\right]^{n-r}$$

**Proof:**

Divide the real line into three intervals:

1. $(-\infty,\; y_r)$
2. $[y_r,\; y_r + h)$ for small $h > 0$
3. $[y_r + h,\; \infty)$

By the **multinomial distribution**, the probability that $r-1$ sample values fall in interval 1, exactly one falls in interval 2, and $n-r$ fall in interval 3 is:

$$\frac{n!}{(r-1)!\,1!\,(n-r)!} \left[\int_{-\infty}^{y_r} f(x)\,dx\right]^{r-1} \left[\int_{y_r}^{y_r+h} f(x)\,dx\right] \left[\int_{y_r+h}^{\infty} f(x)\,dx\right]^{n-r}$$

By the **law of the mean** from calculus:
$$\int_{y_r}^{y_r+h} f(x)\,dx \approx f(y_r)\cdot h$$

Dividing by $h$ and letting $h \to 0$ yields the formula above. $\blacksquare$

---

## Special Cases {#special-cases}

### Smallest Value $Y_1$

$$g_1(y_1) = n\,f(y_1)\left[\int_{y_1}^{\infty} f(x)\,dx\right]^{n-1}$$

### Largest Value $Y_n$

$$g_n(y_n) = n\,f(y_n)\left[\int_{-\infty}^{y_n} f(x)\,dx\right]^{n-1}$$

### Sample Median (odd $n = 2m+1$)

The sample median $\tilde{X} = Y_{m+1}$, with sampling distribution:

$$h(\tilde{x}) = \frac{(2m+1)!}{m!\,m!} \left[\int_{-\infty}^{\tilde{x}} f(x)\,dx\right]^{m} f(\tilde{x}) \left[\int_{\tilde{x}}^{\infty} f(x)\,dx\right]^{m}$$

> For samples of size $n = 2m$, the median is defined as $\dfrac{Y_m + Y_{m+1}}{2}$.

---

## Example 4 – Exponential Population {#example4}

Show that for random samples of size $n$ from an **exponential** population with parameter $\theta$:

$$f(x) = \frac{1}{\theta}e^{-x/\theta}, \quad x > 0$$

the sampling distributions of $Y_1$, $Y_n$, and the median are as derived below.

### Distribution of $Y_1$ (Minimum)

$$\int_{y_1}^{\infty} f(x)\,dx = \int_{y_1}^{\infty} \frac{1}{\theta}e^{-x/\theta}\,dx = e^{-y_1/\theta}$$

Substituting into the formula for $g_1$:

$$g_1(y_1) = n \cdot \frac{1}{\theta}e^{-y_1/\theta} \cdot \left(e^{-y_1/\theta}\right)^{n-1} = \frac{n}{\theta}e^{-ny_1/\theta}$$

$$\boxed{g_1(y_1) = \frac{n}{\theta}\,e^{-ny_1/\theta}, \quad y_1 > 0}$$

This is an **exponential distribution** with parameter $\theta/n$.

---

### Distribution of $Y_n$ (Maximum)

$$\int_{0}^{y_n} f(x)\,dx = 1 - e^{-y_n/\theta}$$

Substituting into the formula for $g_n$:

$$g_n(y_n) = n \cdot \frac{1}{\theta}e^{-y_n/\theta} \cdot \left(1 - e^{-y_n/\theta}\right)^{n-1}$$

$$\boxed{g_n(y_n) = \frac{n}{\theta}\,e^{-y_n/\theta}\!\left(1 - e^{-y_n/\theta}\right)^{n-1}, \quad y_n > 0}$$

---

### Distribution of the Median ($n = 2m+1$)

Computing the required integrals:

$$\int_{-\infty}^{\tilde{x}} f(x)\,dx = 1 - e^{-\tilde{x}/\theta}, \qquad \int_{\tilde{x}}^{\infty} f(x)\,dx = e^{-\tilde{x}/\theta}$$

Substituting:

$$h(\tilde{x}) = \frac{(2m+1)!}{m!\,m!} \cdot \left(1 - e^{-\tilde{x}/\theta}\right)^m \cdot \frac{1}{\theta}e^{-\tilde{x}/\theta} \cdot \left(e^{-\tilde{x}/\theta}\right)^m$$

$$\boxed{h(\tilde{x}) = \frac{(2m+1)!}{m!\,m!}\,\frac{1}{\theta}\,e^{-(m+1)\tilde{x}/\theta}\!\left(1 - e^{-\tilde{x}/\theta}\right)^m, \quad \tilde{x} > 0}$$

---

## Theorem 15 – Asymptotic Distribution of the Median {#theorem15}

For large $n$, the sampling distribution of the median for random samples of size $2n+1$ is approximately **normal** with:

$$\text{Mean} = \tilde{\mu}, \qquad \text{Variance} = \frac{1}{8n\,[f(\tilde{\mu})]^2}$$

where $\tilde{\mu}$ is the population median satisfying $\displaystyle\int_{-\infty}^{\tilde{\mu}} f(x)\,dx = \tfrac{1}{2}$.

**Comparison with the Mean (Normal Population):**

For samples of size $2n+1$ from a normal population, $\tilde{\mu} = \mu$ and $f(\tilde{\mu}) = \dfrac{1}{\sigma\sqrt{2\pi}}$, giving:

| Statistic | Variance |
|:---|:---:|
| Mean $\bar{X}$ | $\dfrac{\sigma^2}{2n+1}$ |
| Median $\tilde{X}$ | $\dfrac{\pi\sigma^2}{4n} \approx \dfrac{\sigma^2}{2n} \cdot \dfrac{\pi}{2}$ |

> **Conclusion:** For large samples from normal populations, the **mean is more reliable** than the median — it has smaller variance (less sampling variability).

---

## Exercises – Week 10 {#exercises-week10}

1. Find the sampling distributions of $Y_1$ and $Y_n$ for random samples of size $n$ from a **continuous uniform** population on $[0, 1]$.

2. Find the sampling distribution of the median for random samples of size $2m+1$ from the population of Exercise 1.
$$h(\tilde{x}) = \frac{(2m+1)!}{m!\,m!}\,\tilde{x}^m(1-\tilde{x})^m, \quad 0 < \tilde{x} < 1$$

3. Find the **mean** and **variance** of $Y_1$ for random samples of size $n$ from the population of Exercise 1.

4. Find the sampling distributions of $Y_1$ and $Y_n$ for random samples of size $n$ from a **Beta distribution** with $\alpha = 3$, $\beta = 2$.
$$g_1(y_1) = 12n\,y_1^2(1-y_1)(1 - 4y_1^3 + 3y_1^4)^{n-1}, \quad 0 < y_1 < 1$$

5. Find the sampling distribution of the median for random samples of size $2m+1$ from the population of Exercise 4.

6. Show that the **joint density** of $Y_1$ and $Y_n$ is:
$$g(y_1, y_n) = n(n-1)\,f(y_1)\,f(y_n)\left[\int_{y_1}^{y_n} f(x)\,dx\right]^{n-2}, \quad y_1 < y_n$$

   **(a)** Use this to find the joint density of $Y_1$ and $Y_n$ for an exponential population.

   **(b)** Use this to find the joint density of $Y_1$ and $Y_n$ for the uniform population of Exercise 1.

7. Using part (b) of Exercise 6, find $\text{Cov}(Y_1, Y_n)$.
$$\text{Cov}(Y_1, Y_n) = \frac{1}{(n+1)^2(n+2)}$$

8. Use the joint density of $Y_1$ and $Y_n$ from Exercise 6 and the **transformation technique** to find an expression for the joint density of $Y_1$ and the sample range $R = Y_n - Y_1$.

9. Using Exercise 8 and part (a) of Exercise 6, find the sampling distribution of $R$ for random samples of size $n$ from an **exponential population**.
$$f(R) = \frac{n(n-1)}{\theta^2}\,e^{-R/\theta}\!\left(1 - e^{-R/\theta}\right)^{n-2} \cdot \frac{1}{n-1}, \quad R > 0$$

10. Use Exercise 8 to find the sampling distribution of $R$ for random samples of size $n$ from the **continuous uniform** population of Exercise 1.

11. Use the result of Exercise 10 to find the **mean** and **variance** of $R$:
$$E(R) = \frac{n-1}{n+1}, \qquad \text{Var}(R) = \frac{2(n-1)}{(n+1)^2(n+2)}$$

---

# Week 11: Characteristic Functions {#week11}

## Introduction {#intro-cf}

In more advanced settings, the **moment-generating function** (MGF) is not always used because many distributions do not have MGFs. Instead, we use the **characteristic function**, which exists for *every* distribution.

> Let $i$ denote the imaginary unit and $t$ an arbitrary real number.

### Definition – Characteristic Function

If $X$ is a random variable, its **characteristic function** is $\varphi_X : \mathbb{R} \to \mathbb{C}$, defined as:

$$\varphi_X(t) = E\!\left(e^{itX}\right) = \begin{cases} \displaystyle\sum_x e^{itx} P(X = x) & \text{if } X \text{ is discrete} \\[8pt] \displaystyle\int_{-\infty}^{\infty} e^{itx} f(x)\,dx & \text{if } X \text{ is continuous} \end{cases}$$

This expectation **exists for all $t$** (since $|e^{itx}| = 1$).

---

## Properties of Characteristic Functions {#properties}

### 1. Uniqueness

$X$ and $Y$ have the same characteristic function **if and only if** they have the same distribution function.

### 2. Moment Generation

If $E(X^n)$ exists, then the $n^{\text{th}}$ derivative of $\varphi_X(t)$ exists and:

$$\varphi_X^{(n)}(t) = i^n\,E\!\left(X^n e^{itX}\right)$$

In particular, at $t = 0$:

$$E(X^k) = \frac{\varphi_X^{(k)}(0)}{i^k}$$

**Proof:**

$$\frac{d}{dt}\varphi_X(t) = \frac{d}{dt}E\!\left(e^{itX}\right) = E\!\left(\frac{d}{dt}e^{itX}\right) = E\!\left(iX e^{itX}\right) = iE\!\left(Xe^{itX}\right)$$

$$\frac{d^2}{dt^2}\varphi_X(t) = i\frac{d}{dt}E\!\left(Xe^{itX}\right) = i\,E\!\left(iX^2 e^{itX}\right) = i^2 E\!\left(X^2 e^{itX}\right)$$

By induction, $\varphi_X^{(n)}(t) = i^n E(X^n e^{itX})$. $\blacksquare$

### 3. Taylor Expansion

If $E(X^n)$ exists:

$$\varphi_X(t) = \sum_{j=0}^{n} \frac{(it)^j}{j!} E(X^j) + o_n(t)$$

where $o_n(t)$ satisfies $\displaystyle\lim_{t \to 0} \dfrac{o_n(t)}{t^n} = 0$.

---

## Mean and Variance via Characteristic Functions {#mean-variance-cf}

From Property 2, evaluating derivatives at $t = 0$:

$$\boxed{E(X) = \frac{\varphi_X'(0)}{i}}$$

$$\boxed{\text{Var}(X) = \frac{\varphi_X''(0)}{i^2} - \left[\frac{\varphi_X'(0)}{i}\right]^2 = -\varphi_X''(0) - \left[E(X)\right]^2}$$

> **Connection to MGF:** For distributions that have an MGF $M_X(t)$, we have $\varphi_X(t) = M_X(it)$.

---

## Example 1 – Binomial Distribution {#example1-binomial}

Let $X \sim \text{Binomial}(n, p)$ with $q = 1 - p$.

### (a) Characteristic Function

$$\varphi_X(t) = E\!\left(e^{itX}\right) = \sum_{x=0}^{n} e^{itx}\binom{n}{x}p^x q^{n-x}$$

$$= \sum_{x=0}^{n} \binom{n}{x}(pe^{it})^x q^{n-x} = \left(q + pe^{it}\right)^n$$

by the **Binomial Theorem** with $a = q$ and $b = pe^{it}$.

$$\boxed{\varphi_X(t) = \left(q + pe^{it}\right)^n}$$

### (b) Mean and Variance

**First derivative:**

$$\varphi_X'(t) = n\left(q + pe^{it}\right)^{n-1} \cdot pi e^{it}$$

At $t = 0$, $q + p = 1$:

$$\varphi_X'(0) = npi \implies E(X) = \frac{npi}{i} = np$$

**Second derivative** (using the product rule):

$$\varphi_X''(t) = n(n-1)\left(q + pe^{it}\right)^{n-2}(pie^{it})^2 + n\left(q + pe^{it}\right)^{n-1} \cdot pi^2 e^{it}$$

At $t = 0$:

$$\varphi_X''(0) = n(n-1)p^2 i^2 + npi^2 = -n(n-1)p^2 - np$$

**Variance:**

$$\text{Var}(X) = \frac{\varphi_X''(0)}{i^2} - [E(X)]^2 = \frac{-n(n-1)p^2 - np}{-1} - n^2p^2$$

$$= n(n-1)p^2 + np - n^2p^2 = np - np^2 = np(1-p) = npq$$

$$\boxed{E(X) = np, \qquad \text{Var}(X) = npq}$$

---

## Example 2 – Exponential Distribution {#example2-exponential}

Let $X \sim \text{Exponential}(\theta)$ with $f(x) = \dfrac{1}{\theta}e^{-x/\theta}$, $x > 0$.

### (a) Characteristic Function

$$\varphi_X(t) = \int_0^{\infty} e^{itx} \cdot \frac{1}{\theta} e^{-x/\theta}\,dx = \frac{1}{\theta}\int_0^{\infty} e^{-x(1/\theta - it)}\,dx$$

$$= \frac{1}{\theta} \cdot \frac{1}{1/\theta - it} = \frac{1}{1 - i\theta t}$$

$$\boxed{\varphi_X(t) = \frac{1}{1 - i\theta t}}$$

### (b) Mean and Variance

**First derivative:**

$$\varphi_X'(t) = \frac{i\theta}{(1-i\theta t)^2}$$

At $t = 0$: $\varphi_X'(0) = i\theta$, so:

$$E(X) = \frac{i\theta}{i} = \theta$$

**Second derivative:**

$$\varphi_X''(t) = \frac{2(i\theta)^2}{(1-i\theta t)^3} = \frac{-2\theta^2}{(1-i\theta t)^3}$$

At $t = 0$: $\varphi_X''(0) = -2\theta^2$, so:

$$\text{Var}(X) = \frac{-2\theta^2}{i^2} - \theta^2 = 2\theta^2 - \theta^2 = \theta^2$$

$$\boxed{E(X) = \theta, \qquad \text{Var}(X) = \theta^2}$$

---

## Exercises – Week 11 {#exercises-week11}

For each distribution below, **(a)** find the characteristic function, and **(b)** derive the mean and variance.

**1. Discrete Uniform** — $P(X = x) = \dfrac{1}{N}$ for $x = 0, 1, \ldots, N-1$.

$$\varphi_X(t) = \frac{1}{N} \cdot \frac{1 - e^{iNt}}{1 - e^{it}}, \qquad E(X) = \frac{N-1}{2}, \qquad \text{Var}(X) = \frac{N^2-1}{12}$$

**2. Poisson** — $P(X = x) = \dfrac{e^{-\lambda}\lambda^x}{x!}$ for $x = 0, 1, 2, \ldots$

$$\varphi_X(t) = e^{\lambda(e^{it}-1)}, \qquad E(X) = \lambda, \qquad \text{Var}(X) = \lambda$$

**3. Continuous Uniform** — $f(x) = \dfrac{1}{b-a}$ for $a < x < b$.

$$\varphi_X(t) = \frac{e^{ibt} - e^{iat}}{it(b-a)}, \qquad E(X) = \frac{a+b}{2}, \qquad \text{Var}(X) = \frac{(b-a)^2}{12}$$

**4. Chi-Squared** — $X \sim \chi^2_n$.

$$\varphi_X(t) = (1 - 2it)^{-n/2}, \qquad E(X) = n, \qquad \text{Var}(X) = 2n$$

**5. Normal** — $X \sim N(\mu, \sigma^2)$.

$$\varphi_X(t) = e^{i\mu t - \sigma^2 t^2/2}, \qquad E(X) = \mu, \qquad \text{Var}(X) = \sigma^2$$

---

*End of STA227 Weeks 10–11 Notes*



